% paper/latex/main.tex
% Note: CJK pattern internal keys appear in the pattern glossary (Section 7.4).
% They are represented here using their English transliterations, matching the
% existing skeleton. The original CJK keys are documented in the open-source
% repository under backend/intelligence/pattern_i18n.py.
\documentclass{llncs}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{amsmath,amssymb}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds}
\usepackage{xcolor}
\usepackage{hyperref}

\title{An Ontology-Guided, Explainable Reasoning Framework\\
for Geopolitical Trajectory Analysis}
\author{Anonymous Author(s)}
\institute{Anonymous Institution\\
\email{anonymous@review.blind}\\
\url{https://anonymous.4open.science/r/El-druin-BLIND}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We present \textbf{EL-DRUIN}, an ontology-guided reasoning system for geopolitical
trajectory analysis that combines a typed pattern registry, rule-based composition
semantics, vector-space compatibility scoring, provenance-preserving posterior
computation, and constrained LLM explanation generation. The 21 registered patterns
are organized into a finite algebraic transition structure with a $7 \times 3$
domain--mechanism factorization, motivated by a seven-domain geopolitical ontology
(geopolitics, economics, technology, military, information, legal, social) and three
cross-domain mechanism classes (coercive, cooperative, structural). This factorization
provides a canonical address for each pattern and supports structured confidence
calibration. We explore a non-Archimedean confidence calibration scheme, using a
7-based valuation motivated by the seven-domain partition, which produces
domain-boundary-sensitive confidence updates rather than uniform geometric decay.
Pattern similarity is scored via an 8-dimensional Lie algebra approximation; a
constrained LLM renders all pre-computed numerical outputs into natural-language
explanations without modifying any numerical field. We demonstrate the framework on
fourteen geopolitical news scenarios and compare against an LLM-only baseline,
showing that EL-DRUIN produces fully traceable, deterministic, auditable assessments
with complete provenance chains. The system is fully open-source; the repository link is anonymised for blind review.
\end{abstract}

\textbf{Keywords:} ontology engineering; ontology-guided reasoning; explainable AI; knowledge graphs; geopolitical trajectory analysis; algebraic transition structure; non-Archimedean confidence; neuro-symbolic AI; provenance-preserving inference; p-adic valuation

\section{Introduction}

Large language models have achieved impressive performance on political
text summarisation and question answering. While LLMs are highly effective
for summarization and linguistic interpretation, their confidence estimates
are typically not grounded in explicit symbolic assumptions, provenance
chains, or domain ontologies~\cite{pan2023unifying}. A model asked to assess the
long-run trajectory of US-China technology competition will produce output
shaped by its training distribution rather than from a principled traversal
of an explicit causal or ontological structure.

This gap has practical epistemic consequences. When an LLM produces a
confidence score for a geopolitical assessment, that number lacks an
auditable derivation: it cannot be traced to specific structural assumptions,
ontology-defined priors, or algebraic constraints. EL-DRUIN addresses this
gap by combining formal ontology with symbolic inference, retaining the LLM
only as a \emph{constrained natural-language renderer} of pre-computed
structured outputs.

The key contributions of this paper are:
\begin{enumerate}
  \item \textbf{Ontology-guided reasoning framework}: A typed pattern registry with
    $7 \times 3$ domain--mechanism factorization that enables compositional
    symbolic inference for geopolitical trajectory analysis, without requiring
    labelled training data.
  \item \textbf{Hybrid reasoning pipeline}: A five-stage pipeline combining
    ontology-typed pattern activation, rule-based composition, 8-dimensional Lie
    algebra similarity scoring, and provenance-preserving Bayesian posterior
    computation.
  \item \textbf{Constrained LLM verbalization}: A mechanism that structurally
    separates numerical reasoning from natural-language explanation: the LLM
    receives all numerical fields as locked inputs and is prohibited from modifying
    them; if the LLM call fails, a template fallback preserves all numerical outputs.
  \item \textbf{Open-source implementation and empirical evaluation}: A publicly
    available system with full computation traces, evaluated on fourteen geopolitical
    news scenarios with an LLM-only baseline and ablation study.
\end{enumerate}

The remainder of the paper is organised as follows. Section~2 provides
background on relevant ontology and algebra formalisms. Section~3
describes the system architecture. Sections~4 and~5 present the
mathematical formalisation. Section~6 describes the full five-stage
reasoning pipeline. Section~7 presents experiments. Section~8 discusses
related work and Section~9 concludes.

\section{Background}

\subsection{Ontology Engineering for Event Reasoning}

Formal ontologies represent domain knowledge as typed entities,
relations, and constraints. The CAMEO event coding system~\cite{schrodt2012cameo}
provides a controlled vocabulary of political event types widely used in
computational political science. FIBO (Financial Industry Business
Ontology) provides analogous structure for financial relationships.
EL-DRUIN draws on both, mapping typed entity pairs through relation
predicates to named patterns registered in a Cartesian product registry:
$E \times R \times E \to \mathrm{Pattern}$, where $E$ is the set of
entity types and $R$ is the set of relation types.

\subsection{Finite Semigroups and Fixed Points}

A \textbf{semigroup} $(S, \cdot)$ is a set with an associative binary operation.
A \textbf{finite semigroup} is one where $|S|$ is finite. An element $e \in S$
is \textbf{idempotent} if $e \cdot e = e$. It is a classical
result~\cite{green1951structure} that every finite semigroup contains idempotent
elements, and that iteration of the operation from any starting state
eventually reaches a set of idempotents. These idempotents are the
\textbf{attractors} of the system---they correspond to states from which no
further composition moves the system to a genuinely new state.

In EL-DRUIN, the set $S$ is the set of named Dynamic Patterns, and the
partial binary operation is defined by
$\mathit{composition\_table}[(A, B)] = C$, read as ``when pattern $A$ and
pattern $B$ are simultaneously active, the system tends toward pattern
$C$''. This table is validated for algebraic closure (all outputs $C$
must be known patterns) and inverse consistency (if $\mathit{inverse}(A) = B$
then $\mathit{inverse}(B) = A$) at application startup.

\subsection{Lie Algebra Approximation}

A \textbf{Lie algebra} $\mathfrak{g}$ over a field $F$ is a vector space equipped
with a bilinear antisymmetric bracket
$[\cdot, \cdot] \colon \mathfrak{g} \times \mathfrak{g} \to \mathfrak{g}$
satisfying the Jacobi identity. Lie algebras arise as the tangent space
of Lie groups at the identity and encode the infinitesimal structure of
continuous symmetry transformations.

In EL-DRUIN, we do not work with a formal Lie group; instead we
construct a \textbf{finite-dimensional approximation}: each named pattern $P$
is assigned a vector $v_P \in \mathbb{R}^8$ encoding its intensity across
eight semantic dimensions (coercion, cooperation, dependency,
information, regulation, military, economic, technology). Pattern
composition $A \oplus B$ is approximated by vector addition $v_A + v_B$,
and the resulting target pattern is identified by maximising cosine
similarity $\cos(v_A + v_B, v_C)$ over all known patterns $C$.

Each named pattern $P$ is embedded as a skew-symmetric matrix
$X_P \in \mathfrak{so}(8)$ via the \emph{hat map}
\begin{equation}
  \hat{v}[i,j] = v[i] - v[j], \quad \hat{v}[i,i] = 0,
  \label{eq:hatmap}
\end{equation}
equivalently $X_P = v_P \mathbf{1}^\top - \mathbf{1} v_P^\top$, which is
antisymmetric by construction ($X_P^\top = -X_P$)~\cite{hall2015}.
The \textbf{Lie bracket} is the matrix commutator
\begin{equation}
  [X_A, X_B] = X_A X_B - X_B X_A \in \mathfrak{so}(8),
  \label{eq:liebracket}
\end{equation}
which satisfies antisymmetry, bilinearity, and the Jacobi identity
exactly~\cite{humphreys1972,hall2015}. The Frobenius norm
$\|[X_A, X_B]\|_F$ measures non-commutativity: it equals zero if and
only if $X_A$ and $X_B$ commute (i.e., pattern-activation order is
immaterial), and is large when the two orderings produce structurally
distinct outcomes.

\textit{An important structural property}: while each pattern vector $v_P$
maps into the $(n-1)$-dimensional image of the hat map
$\hat{\cdot} \colon \mathbb{R}^8 \to \mathfrak{so}(8)$, the commutator
$[X_A, X_B] = X_A X_B - X_B X_A$ lives in the full $\binom{8}{2}=28$-dimensional
$\mathfrak{so}(8)$, outside the hat-map image. This is not a
defect---it is the feature that makes the Lie-algebra path genuinely
independent of the Bayesian path: the bracket activates semantic
directions unreachable by any single pattern vector, capturing
interaction effects that the additive sum $v_A + v_B$ cannot represent.

\section{System Architecture}

EL-DRUIN is a multi-module backend (FastAPI + KuzuDB) with a Streamlit
frontend. The system has four functional layers:

\begin{itemize}
\item \textbf{Ontology Layer}: \texttt{CARTESIAN\_PATTERN\_REGISTRY} (21 named
  patterns across geopolitics, economics, technology, military, information,
  legal, and social domains), \texttt{composition\_table} (16 composition rules),
  \texttt{inverse\_table} (21 inverse pairs), \texttt{DOMAIN\_MAP} (7-partition
  of patterns by geopolitical domain), \texttt{MECHANISM\_MAP} (3-partition of
  patterns by mechanism class), and \texttt{EntityType} / \texttt{RelationType} enumerations.
\item \textbf{Lie Algebra Layer} (\texttt{lie\_algebra\_space.py}): 8-dimensional
  pattern vectors, \texttt{LieAlgebraSpace.add()}, \texttt{bracket()},
  \texttt{project()}, and \texttt{phase\_detect()} operations, PCA projection
  for 2D visualisation.
\item \textbf{Pipeline Layer} (\texttt{evented\_pipeline.py}): Five-stage reasoning
  pipeline (event extraction $\to$ pattern activation $\to$ transition
  enumeration $\to$ state vector computation $\to$ Bayesian conclusion
  generation).
\item \textbf{Forecasting Layer} (\texttt{ontology\_forecaster.py}): Multi-step
  simulation over the group transition graph, attractor
  detection, p-adic bifurcation point identification, and 7-adic
  confidence calibration (replacing geometric step-decay $\lambda^t$).
\end{itemize}

The frontend exposes five tabs for each analysis: (1) Conclusion with
structured alpha/beta paths; (2) Events extracted by the compound event
rules; (3) Active and derived patterns with full composition logic
chain; (4) Probability Tree with Bayesian compute trace showing the
exact formula
$\mathit{prior}_A \times \mathit{prior}_B \times \mathit{lie\_similarity} / Z$;
(5) Lie Algebra 8D state vector with dimensional breakdown.

\section{Ontology Schema and Pattern Registry}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    node distance=1.4cm and 2.0cm,
    every node/.style={font=\small},
    class/.style={draw, rectangle, rounded corners=4pt, fill=blue!10,
                  minimum width=2.6cm, minimum height=0.7cm, align=center},
    abstract/.style={draw, rectangle, rounded corners=4pt, fill=gray!15,
                     minimum width=2.6cm, minimum height=0.7cm,
                     align=center, dashed},
    rel/.style={-{Stealth[length=6pt]}, thick},
    partof/.style={-{Diamond[open,length=6pt]}, thick, dashed},
    label/.style={font=\footnotesize\itshape, midway},
]

%% Core classes
\node[class] (event)    {Event};
\node[class] (entity)   [below left=of event]   {EntityType};
\node[class] (reltype)  [below right=of event]  {RelationType};
\node[class] (pattern)  [below=2.2cm of event]  {\textbf{DynamicPattern}};
\node[class] (domain)   [below left=of pattern] {Domain (×7)};
\node[class] (mech)     [below right=of pattern]{MechanismClass (×3)};
\node[abstract](traj)   [below=1.8cm of pattern]{Trajectory};
\node[class] (prov)     [right=2.5cm of pattern]{ProvenanceRecord};

%% Relations: Event → EntityType / RelationType
\draw[rel] (event) -- node[label, left]{src/tgt type} (entity);
\draw[rel] (event) -- node[label, right]{rel type}    (reltype);

%% Event → DynamicPattern (grounding)
\draw[rel, bend left=15] (event) to
    node[label, right]{activates} (pattern);

%% DynamicPattern → Domain / MechanismClass (factorization)
\draw[rel] (pattern) -- node[label, left]{domain} (domain);
\draw[rel] (pattern) -- node[label, right]{mechanism} (mech);

%% DynamicPattern → Trajectory (composition chain)
\draw[rel, dashed] (pattern) to
    node[label, right]{composition$^*$} (traj);

%% DynamicPattern → ProvenanceRecord
\draw[partof] (prov) -- node[label, above]{compute\_trace\_ref} (pattern);

%% Ontology-typed confidence prior annotation
\node[font=\footnotesize, draw=none, text=gray, right=0.2cm of mech]
    {\textit{confidence\_prior}};

\end{tikzpicture}
\caption{EL-DRUIN ontology schema. \textbf{DynamicPattern} is the central
mid-level semantic unit bridging event-level data (top) and trajectory-level
reasoning (bottom). Each pattern carries a \emph{domain} assignment (one of
seven geopolitical domains) and a \emph{MechanismClass} assignment (one of
three cross-domain mechanism types), providing a $7 \times 3$ factorization
that grounds compositional inference and confidence calibration in the ontology
structure. The \emph{ProvenanceRecord} links every numerical output to its
derivation via \texttt{compute\_trace\_ref}.}
\label{fig:ontology-schema}
\end{figure}

\subsection{Entity and Relation Type Enumerations}

The entity type set $E$ contains 18 elements spanning geopolitical
(\texttt{STATE}, \texttt{ALLIANCE}, \texttt{PARAMILITARY}, \texttt{IDEOLOGY}),
economic (\texttt{FIRM}, \texttt{FINANCIAL\_ORG}, \texttt{RESOURCE}, \texttt{CURRENCY},
\texttt{SUPPLY\_CHAIN}), technological (\texttt{TECH}, \texttt{STANDARD}),
social-cognitive (\texttt{PERSON}, \texttt{MEDIA}, \texttt{TRUST}, \texttt{INSTITUTION}),
and event types (\texttt{CONFLICT}, \texttt{NORM}), plus the \texttt{UNKNOWN} sentinel
for unresolvable types. The relation type set $R$ contains 20 elements
across four categories: coercive/adversarial (\texttt{SANCTION},
\texttt{MILITARY\_STRIKE}, \texttt{COERCE}, \texttt{BLOCKADE}), cooperative
(\texttt{SUPPORT}, \texttt{ALLY}, \texttt{AID}, \texttt{AGREE}), dependency/flow
(\texttt{DEPENDENCY}, \texttt{TRADE\_FLOW}, \texttt{SUPPLY}, \texttt{FINANCE}), and
structural/institutional (\texttt{SIGNAL}, \texttt{PROPAGANDA}, \texttt{LEGITIMIZE},
\texttt{DELEGITIMIZE}, \texttt{REGULATE}, \texttt{STANDARDIZE}, \texttt{EXCLUDE},
\texttt{INTEGRATE}).

\subsection{Dynamic Pattern Registration}

Each pattern $P \in S$ is registered by a call
$\mathit{\_reg}(e_{\mathrm{src}},\allowbreak\; r,\allowbreak\; e_{\mathrm{tgt}},\allowbreak\;
\mathit{pattern\_name},\allowbreak\; \mathit{domain},\allowbreak\;
\mathit{typical\_outcomes},\allowbreak\; \mathit{mechanism\_class},\allowbreak\;
\mathit{inverse\_pattern},\allowbreak\; \mathit{composition\_hints},\allowbreak\;
\mathit{confidence\_prior})$.
The pattern name is a human-readable string (e.g., ``Hegemonic
Sanctions'') to facilitate integration with geopolitical analysis
workflows. Table~\ref{tab:patterns} summarises 10 representative entries
from the 21 registered patterns.

\begin{table}[ht]
\centering
\caption{Selected Dynamic Patterns (10 of 21 shown). Full registry
  available in the open-source repository.}
\label{tab:patterns}
\begin{tabular}{@{}p{0.26\linewidth}p{0.11\linewidth}p{0.20\linewidth}
                   p{0.07\linewidth}p{0.24\linewidth}@{}}
\toprule
Pattern Name & Domain & Mechanism Class & Prior & Inverse Pattern \\
\midrule
Hegemonic Sanctions & geopolitics & coercive\_leverage & 0.78
  & Sanctions Relief / Normalisation \\
Entity-List Technology Blockade & technology & tech\_denial & 0.80
  & Technology Licence / Unblocking \\
Interstate Military Conflict & military & kinetic\_escalation & 0.75
  & Ceasefire / Peace Agreement \\
Great-Power Coercion/Deterrence & geopolitics & coercive\_leverage & 0.70
  & Diplomatic Concession / De-escalation \\
Multilateral Alliance Sanctions & geopolitics & multilateral\_pressure & 0.73
  & Multilateral Sanctions Relief \\
Tech Decoupling / Technology Iron Curtain & technology & tech\_decoupling & 0.76
  & Technology Cooperation Reintegration \\
Financial Isolation / SWIFT Cut-Off & economics & financial\_exclusion & 0.79
  & Financial Reintegration \\
Bilateral Trade Dependency & economics & economic\_interdependence & 0.71
  & Trade War / Decoupling \\
Technology Standards Leadership & technology & tech\_governance & 0.72
  & Standard Competition Failure \\
Information Warfare / Narrative Control & information & epistemic\_warfare & 0.68
  & Information Environment Restoration \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Algebraic Consistency Validation}

Two static validators are executed at application startup:

\textbf{validate\_inverses()}: Checks that for every $(A, B)$ pair in
\texttt{inverse\_table}, the reverse pair $(B, A)$ is also present. This
enforces the group-theoretic condition that the inverse of an inverse
returns the original element.

\textbf{validate\_composition\_closure()}: Checks that for every
$(A, B) \to C$ in \texttt{composition\_table}, $C$ is a known pattern name in
the registry. This enforces algebraic closure---the composition of two
patterns must remain within the pattern set, preventing the system from
producing unregistered states. A coupling-asymmetry metric is computed
in the frontend for each active analysis to measure the concentration of
active mechanism classes across semantic domains.\footnote{The frontend
coupling-asymmetry metric uses the Herfindahl--Hirschman Index
normalised against a fixed domain baseline of $n_d = 8$ (the total
number of semantic domains in the ontology), not against the count of
currently active domains, to avoid division by zero when a single domain
is active.}

\subsection{Ontology Design Pattern: Trajectory Reasoning}
\label{sec:design-pattern}

Figure~\ref{fig:ontology-schema} illustrates the central ontology design
decision in EL-DRUIN. Rather than merely enumerating entities and relations,
we introduce a \textbf{trajectory reasoning design pattern} in which
\emph{DynamicPattern} objects serve as typed mid-level semantic units that
bridge event-level observations and trajectory-level reasoning.

\paragraph{Why this is more than a pattern list.}
Existing political event coding schemes (CAMEO, GDELT, ICEWS) register events
as typed actor-action-target triples and aggregate them statistically. EL-DRUIN
differs in three ways:
\begin{enumerate}
  \item \textbf{Composition semantics.} Each \texttt{DynamicPattern} participates
    in a composition table that defines how pairs of active patterns interact to
    produce derived patterns. This elevates patterns from database entries to
    active inference nodes in a symbolic reasoning graph.
  \item \textbf{Ontology-grounded confidence.} Each pattern carries a
    \texttt{confidence\_prior} that is an ontology-defined belief about the
    base rate of that pattern type, not a learned parameter or an arbitrary
    constant. The prior participates directly in the Bayesian posterior
    computation, making every confidence value traceable to an explicit
    ontological assumption.
  \item \textbf{Provenance-preserving structure.} Every numerical output is
    anchored to a \texttt{ProvenanceRecord} via \texttt{compute\_trace\_ref}.
    If the system changes a confidence value, the record changes too. This
    satisfies the ISWC community's standard for auditable and reproducible
    ontological inference.
\end{enumerate}

\paragraph{Why LLM is needed, and where.}
The LLM plays a strictly bounded role in EL-DRUIN:
\begin{itemize}
  \item \textbf{Not responsible for}: numerical reasoning, confidence computation,
    pattern activation, composition, or attractor detection. These are entirely
    determined by the ontology and the symbolic pipeline.
  \item \textbf{Responsible for}: rendering the pre-computed structured output
    into natural-language strategic interpretation (2--3 sentences).
  \item \textbf{Locked fields}: the LLM prompt presents all numerical fields as
    locked values that must not be changed. Any output that modifies a numerical
    field, introduces an entity not present in the structured output, or produces
    CJK characters triggers a guardrail fallback to template-based rendering.
  \item \textbf{Failure isolation}: if the LLM call fails entirely, a template
    fallback produces the \texttt{conclusion.text} field using pre-computed
    driving-factor statements. No numerical output is degraded.
\end{itemize}

This architecture is a deliberate design choice: it ensures that the system's
\emph{analytical conclusions} --- the patterns, transitions, posteriors, and
attractor predictions --- are fully reproducible and auditable even in the absence
of any LLM, while the LLM provides the linguistic fluency that makes those
conclusions accessible to non-specialist users.

\paragraph{The $7 \times 3$ factorization as reasoning semantics.}
The \texttt{domain} $\times$ \texttt{mechanism\_class} factorization is not an
external label applied post-hoc; it is a structural design choice with reasoning
consequences. Patterns in the same domain share a confidence-calibration
baseline (the 7-based valuation period), while patterns of the same mechanism
class tend to compose into predictable transition chains (coercive patterns
escalate into sanctions or military action; cooperative patterns de-escalate into
normalization). This factorization makes cross-domain and cross-mechanism
propagation explicit, which is essential for explainability: when the system
predicts a cross-domain transition (e.g., from economic sanctions to military
coercion), the analyst can inspect exactly which composition rules fired and why.

\section{Mathematical Formalisation}

\noindent\textbf{Note on mathematical structure.}
Throughout this section, we borrow organizational principles from three mathematical
structures---finite semigroup theory, Lie algebras over $\mathfrak{so}(8)$, and
$p$-adic valuations---without claiming that our construction satisfies the complete
axiomatic systems of these theories. Specifically: (1) we use semigroup idempotent
theory as a \emph{termination criterion} for forward simulation, not as a claim that
our partial composition table forms a full semigroup; (2) we embed patterns in
$\mathfrak{so}(8)$ to compute a \emph{directional similarity score} (cosine of the
vector sum), not to perform Lie group integration or claim differential-geometric
properties; (3) we use the 7-adic absolute value as a \emph{domain-boundary-sensitive
decay function} calibrated to the seven-domain ontology partition, not as a claim of
formal $p$-adic geometry over $\mathbb{Q}_7$. We use the phrase
``design-motivated formalization'' throughout to signal that a mathematical structure
is being used for its organizational properties rather than its full theorem apparatus.

\subsection{Algebraic Organization of the Pattern Space}
\label{sec:pattern-algebra}

We organize the 21 registered patterns into a \textbf{finite algebraic transition
structure} with a $7 \times 3$ domain--mechanism factorization. This factorization
is \emph{motivated by} the structure of groups of order $21 = 3 \times 7$, and we
use the corresponding Sylow-inspired partition as a \emph{design heuristic} rather
than claiming that the composition table satisfies full group axioms: the current
composition table contains 16 defined rules, covering approximately $3.6\%$ of all
$21 \times 21$ pattern pairs, and full Cayley table completion is future work
(Section~9).

Let $S = \{p_1, \ldots, p_{21}\}$ be the set of 21 registered patterns. The
\textbf{domain partition} assigns each pattern to one of seven geopolitical domains
(geopolitics, economics, technology, military, information, legal, social),
forming seven groups of size 3. The \textbf{mechanism partition} assigns each
pattern to one of three cross-domain mechanism classes (coercive, cooperative,
structural), forming three groups of size 7. Together, these two partitions
provide a \textbf{canonical domain $\times$ mechanism address} for each pattern,
supporting structured confidence calibration and ontology-guided traversal.

This $7 \times 3$ factorization is inspired by the algebraic structure of groups
of order $21 = 3 \times 7$ (specifically, the semidirect product $\mathbb{Z}_7
\rtimes \mathbb{Z}_3$), in which the order-7 and order-3 subgroups partition
elements in exactly this way. We use this structure as an \emph{organizational
scheme}, not as a claim that $(S, \cdot)$ with the partial composition table
satisfies all group axioms.

We also retain the original power-set operation for forward simulation:
define the partial binary operation
${\cdot} \colon S \times S \to S \cup \{\bot\}$ by:
\[
  p_i \cdot p_j =
  \begin{cases}
    \mathit{composition\_table}[(p_i, p_j)]
      & \text{if } (p_i, p_j) \in \operatorname{dom}(\mathit{composition\_table}) \\
    \bot & \text{otherwise.}
  \end{cases}
\]

We extend $\cdot$ to a total operation on the power set $\mathcal{P}(S)$
by:
\[
  A \cdot B = \{ c \in S : \exists\, a \in A,\, b \in B
               \text{ s.t.\ } a \cdot b = c \}
           \cup \{ a \in A : \mathit{inverse\_table}[a] \in A^c \}
\]
(the second term accounts for low-probability inverse transitions). The
forward simulation proceeds by iterating this operation from the initial
pattern set $S_0$; a step is productive whenever
$|S_{t+1} \setminus S_t| > 0$, i.e.\ when at least one genuinely new
pattern enters the reachable set. Note that $(\mathcal{P}(S), \cdot)$ is itself
a semigroup (the semigroup of subsets under the induced operation).

\subsection{Lie Algebra Embedding}

Each pattern $p_i$ is assigned a vector $\varphi(p_i) = v_i \in
\mathbb{R}^8$ by manual annotation across eight semantic dimensions
$d = (\text{coercion, cooperation, dependency, information, regulation,
military, economic, technology})$, with $v_i[d] \in [-1, 1]$.
The \textbf{Lie bracket} is the matrix commutator on $\mathfrak{so}(8)$:
\begin{equation}
  [X_A, X_B] = X_A X_B - X_B X_A,
  \label{eq:commutator52}
\end{equation}
where $X_P = \hat{v}_P \in \mathfrak{so}(8)$ is the hat-map embedding
of Equation~\eqref{eq:hatmap}~\cite{humphreys1972}.

The \textbf{Lie similarity} of composition step $(A, B) \to C$ is defined as:
\[
  \mathit{lie\_sim}(A, B, C)
    = \cos(v_A + v_B,\, v_C)
    = \frac{\langle v_A + v_B,\, v_C \rangle}{\|v_A + v_B\| \cdot \|v_C\|}
\]

This quantity measures how well the vector sum of the two composing
patterns approximates the target pattern in direction, providing a
continuous-valued plausibility score for each discrete composition rule.

\textit{Path independence diagnostic.} The quantity
$\delta = 1 - |\cos(v_A + v_B, \eta)| \in [0, 1]$, where
$\eta = \operatorname{row\_norms}([X_A, X_B])$, measures the structural
independence of the two inference paths. $\delta \approx 0$ indicates
the Lie bracket interaction reinforces the same semantic direction as
the additive sum (linear regime---the two paths are redundant);
$\delta \approx 1$ indicates the bracket activates orthogonal dimensions
absent from $v_A + v_B$ (nonlinear emergence regime). This diagnostic is
the operational definition of ``structural novelty beyond additive
inference'': a high-$\delta$ result is the strongest evidence that the
dual-path architecture provides information not available to a purely
Bayesian system.

\subsection{p-Adic Confidence and Revised Bayesian Posterior}
\label{sec:padic}

We replace the geometric step-decay $\lambda^t$ (where $\lambda = 0.85$ was
an ad hoc constant) with a 7-based non-Archimedean calibration motivated by
the seven-domain ontology partition. Selecting $p = 7$ to match the number of
domain classes imposes \emph{domain-boundary-sensitive} confidence updates:
confidence is stable within a domain-aligned reasoning horizon and drops sharply
at multiples of 7, where a cross-domain phase transition is structurally
plausible. This is a design-motivated choice rather than a claim of formal
p-adic geometry; we use the 7-adic absolute value as a principled
non-Archimedean alternative to the arbitrary constant $\lambda = 0.85$:
\[
  c(P, t) = c_0^{(P)} \cdot |t|_7 = c_0^{(P)} \cdot 7^{-v_7(t)}
\]
where $v_7(t)$ is the 7-adic valuation of $t$ (the largest $k$ such that
$7^k \mid t$). At $t \notin 7\mathbb{Z}$, $|t|_7 = 1$ and confidence equals
the base prior. At $t \in \{7, 14, 21, \ldots\}$, $|t|_7 \leq \tfrac{1}{7}$,
marking \emph{domain-level phase transitions} where the system reassesses its
regime estimate.

The revised \textbf{posterior weight} of transition $(A, B) \to C$ at step $t$ is:
\[
  w_t(A, B \to C)
    = \pi_t(A) \cdot \pi_t(B) \cdot \mathit{lie\_sim}(A, B, C) \cdot |t|_p
\]
replacing the previous $\lambda^t$ factor. This provides a principled,
domain-sensitive alternative to the arbitrary decay constant, with confidence
updates sharpening at multiples of 7 where cross-domain phase transitions are
structurally plausible.

The \textbf{aggregate confidence} of the forecast is:
\[
  c_{\mathrm{final}} = c_0 \cdot |T|_7,
  \quad
  c_0 = \operatorname{mean}(\{\mathit{confidence\_prior}(p) : p \in S_0\})
\]
This formulation makes the provenance of every probability fully
transparent: $c_{\mathrm{final}}$ is a product of the ontology-defined prior
and the 7-adic absolute value of the horizon step count.

\textit{Path integration.} Where the dual-path architecture is active,
the final posterior is adjusted by a path-consistency term:
\[
  \hat{w}(C)
    = P_{\mathrm{Bayes}}(C)
      \cdot \frac{1 + \alpha \cdot \max(0,\, \mathit{consistency})}{1 + \alpha},
  \quad \alpha = 0.30
\]
where $\mathit{consistency} = \cos(\eta, v_C)$ measures the cosine alignment
between the Lie-bracket emergence vector $\eta$ and the target pattern
vector. At $\mathit{consistency} = 0$ (orthogonal paths), the formula applies
a structural 23\% discount ($\hat{w} \approx 0.77 \cdot P_{\mathrm{Bayes}}$),
encoding prior skepticism that resolves only when paths converge. At
$\mathit{consistency} = 1$, confidence is boosted by at most
$\alpha / (1 + \alpha) \approx 23\%$.

\paragraph{Ultrametric distance on pattern space.}
\label{sec:ultrametric}
The Sylow-7 partition induces a 7-adic ultrametric
$d_7(A, B) = 7^{-v_7(\mathrm{graph\_dist}(A,B))}$
on $S$, where $\mathrm{graph\_dist}(A,B)$ is the shortest path length in the
composition graph $\Gamma = (S, E)$. Patterns in the same domain have distance
$7^{-k} \ll 1$ (easy intra-domain propagation); cross-domain patterns have
distance $1$ (maximum, requiring a phase transition). The ultrametric inequality
$d_7(A,C) \leq \max(d_7(A,B), d_7(B,C))$ models the design intuition that
geopolitical crises are either domain-contained or fully cross-domain.
We use this as a design-motivated formalisation, not an empirical claim.

\subsection{Attractor Detection}

A pattern $P \in S_T$ (the pattern set at convergence step $T$) is an
\textbf{attractor} (fixed point of the power-set semigroup) if:
\[
  \forall\, Q \in S_T \colon P \cdot Q \in S_T
  \quad \text{(closure within the current set).}
\]
This condition identifies patterns from which all further composition
operations remain within the already-activated set---the algebraic
termination condition. In the limit $T \to \infty$, these correspond to
the idempotent elements of the finite semigroup.

\subsection{Phase Transition Detection}

A \textbf{phase transition} at step $t$ is flagged when the $L_2$ norm of
the state vector shift exceeds a threshold $\theta$:
\[
  \|v_t - v_{t-1}\|_2 > \theta, \quad \theta = 0.25
\]
(\emph{calibrated on unit-normalised embeddings; for unnormalised vectors
with typical norm $\|v\| \approx 1.5$, the effective angular displacement
is approximately $9.6^\circ$})

\noindent where $v_t = \sum_i \pi_t(p_i) \cdot \varphi(p_i)$ is the
weighted mean vector at step $t$. This condition detects when a
continuous change in pattern weights has caused a discontinuous shift in
the dominant semantic regime---i.e., the system has crossed a phase
boundary in the Lie algebra space.

\subsection{p-Adic Bifurcation Detection}
\label{sec:bifurcation}

The original bifurcation condition (posterior gap $< 0.15$) is replaced by a
structurally motivated p-adic condition:
\[
  \text{bifurcation at step } t
  \iff
  v_p\!\left(\pi_t(p_{\mathrm{top1}}) - \pi_t(p_{\mathrm{top2}})\right) \geq k_0
\]
where $k_0 = 1$ is the minimum valuation threshold and $p = 7$.  A large
valuation means the posterior weight difference is divisible by a high power
of 7, making the two leading attractors \emph{p-adically indistinguishable}.
This provides a domain-structure-aligned notion of near-tie: it detects when
the system is at a bifurcation point \emph{at the granularity of the 7-domain
partition}, complementing the real-valued threshold with a structurally motivated
discrete criterion.

Implementation uses exact rational arithmetic (\texttt{fractions.Fraction}) to
avoid floating-point artefacts in the valuation computation.

\section{The Five-Stage Reasoning Pipeline}

For real-time news analysis, EL-DRUIN operates a five-stage pipeline
(implemented in \texttt{evented\_pipeline.py}):

\textbf{Stage 1: Event Extraction.}
News text is processed by a compound-rule event extractor that maps
keyword co-occurrences to event types. A key design decision is that
compound rules enforce co-activation: a military strike event and a
humanitarian crisis event are both produced when strike-related keywords
and casualty-related keywords co-occur in the same text. This guarantees
$\geq\!2$ active events for military/crisis news, which is necessary to
trigger composition-based derived patterns (a composition requires two
active patterns).

\textbf{Stage 2a: Pattern Activation.}
Each extracted event type is mapped to a set of candidate
$(e_{\mathrm{src}}, r, e_{\mathrm{tgt}})$ triples via a domain hint
table. For each candidate triple, \texttt{lookup\_pattern\_by\_strings()}
retrieves the registered \texttt{DynamicPattern} (with fallback to fuzzy
matching at score $\geq 0.4$). The activated pattern's
\texttt{confidence\_prior} is scaled by the event's extraction confidence.

\textbf{Stage 2b: Transition Enumeration.}
The full \texttt{composition\_table} is scanned for entries where at least
one of the pattern arguments appears in the active set. For each
matching entry $(A, B) \to C$, the posterior weight
$w = \pi(A) \cdot \pi(B) \cdot \mathit{lie\_sim}(A, B, C)$ is computed.
Simultaneously, the \texttt{inverse\_table} is checked for each active
pattern, generating low-probability inverse transitions at weight
$0.20 \times \pi(A)$. All transitions are sorted by posterior weight;
the top-5 are returned as the derived pattern set.

\textit{Note on KG weight fall-back.} When no KG evidence is available
($w_{\mathrm{KG}} = 0$), the trigger amplification formula reduces to
$a = \operatorname{clamp}(w_{\mathrm{causal}} + \Delta_{\mathrm{domain}})$
with $w_{\mathrm{causal}}$ receiving full effective weight; the 0.4 KG
coefficient disappears rather than being redistributed. This is
intentional: the formula is non-normalising by design, bounding the
amplification factor within $[0, 1]$ via the clamp operation.

\textbf{Stage 2c: State Vector Computation.}
The weighted mean vector $v = \sum_i \pi(p_i) \cdot \varphi(p_i)$ is
computed over all active patterns, producing the 8D ontological state
vector. PCA (via \texttt{numpy.linalg.svd}) projects this to 2D for
frontend scatter plot visualisation. The dominant semantic dimension is
identified as $\operatorname{argmax}_d |v[d]|$.

\textbf{Stage 2d: Driving Factor Aggregation.}
Active patterns are grouped by \texttt{mechanism\_class}. For each group,
the weighted sum $\sum \pi(p_i)$ gives a group weight, and the
\texttt{typical\_outcomes} of all patterns in the group are weighted by
their pattern's \texttt{confidence\_prior}. The top-3 outcomes per group
are formatted as human-readable driving factor statements using a
template library keyed by \texttt{mechanism\_class}. This step is entirely
deterministic---no LLM call.

\textbf{Stage 3: Bayesian Conclusion Generation.}
Alpha and beta paths are constructed from the top two transitions by
posterior weight. Composite confidence is computed as
$\operatorname{mean}(\mathit{confidence\_prior}$ over active patterns$)
\times \sqrt{\mathit{verifiability} \times \mathit{KG\_consistency}}$.
The LLM is called only to write the \texttt{conclusion.text} field, with a
prompt that (a) presents all pre-computed fields as locked values,
(b) forbids modification of any numerical field, and (c) requests 2--3
sentences of strategic-language interpretation. If the LLM call fails,
a template fallback produces the text field without degrading any
numerical output.

\section{Evaluation}

\subsection{Dataset and Setup}
\label{sec:methodology}

We evaluate EL-DRUIN on two complementary experimental protocols:

\textbf{(A) News analysis and baseline comparison.}
Fourteen geopolitical news excerpts drawn from publicly accessible sources
(Reuters, BBC News, Financial Times, Brookings Institution, CSIS, NATO, Wall
Street Journal) are processed by the EL-DRUIN pipeline and by a GPT-4o
baseline~\cite{pan2023unifying}. The excerpts span nine thematic categories
(technology export control, trade war, financial isolation, military coercion,
energy weaponisation, military alliance, multilateral sanctions, supply-chain
dependency, information warfare) and cover events from 2018 to 2024. EL-DRUIN
is run in fully deterministic mode ($\sigma = 0$ across repeated runs). GPT-4o
is queried $N = 5$ times per excerpt with identical prompts to measure output
variance. The full dataset, collection protocol, and raw results are available
in \texttt{experiments/news\_samples.jsonl} and
\texttt{experiments/gpt\_manual\_data.py}.

\textbf{(B) Historical backtesting.}
We evaluate trajectory prediction against 8 curated historical events with known
realized outcomes, using the framework in \texttt{experiments/backtest\_evaluator.py}.
For each event, the pipeline is run on a contemporaneous news excerpt (i.e., text
available at the time of the event); the predicted primary attractor is compared
against the curated realized direction. We report direction accuracy (exact match
= 1.0, same ontology domain = 0.5, miss = 0.0), Brier score
$(p - \text{outcome})^2$, attractor identification score, and a composite weighted
score $0.4 \times \text{direction} + 0.3 \times (1 - \text{Brier}) + 0.3 \times
\text{attractor}$.

\textbf{(C) Ablation study.}
We compare four configurations to isolate the contribution of each pipeline
component:
\begin{itemize}
  \item \textbf{Baseline 1 (LLM-only)}: GPT-4o receives the same news excerpt
    and outputs a trajectory judgment directly, without ontology grounding.
  \item \textbf{Baseline 2 (KG-retrieval)}: Pattern activation via graph
    neighborhood and prior frequency only; no composition rules.
  \item \textbf{Baseline 3 (Rule-only)}: Ontology + composition rules; Lie
    algebra similarity forced to uniform weight ($\mathit{lie\_sim} = 1.0$
    for all transitions); no p-adic calibration.
  \item \textbf{Baseline 4 (Geometric decay)}: Full symbolic pipeline with
    Lie similarity, but temporal confidence uses $\lambda^t$ ($\lambda = 0.85$)
    instead of the 7-based non-Archimedean calibration.
  \item \textbf{EL-DRUIN (Full)}: Complete system as described in Sections 4--6.
\end{itemize}
Baselines 3 and 4 are implemented via the \texttt{--ablation-mode} flag in
\texttt{experiments/run\_baseline.py}. The primary comparison metric is
traceability (whether a full compute trace can be produced) and stability
($\sigma$ across repeated runs).

\subsection{Results}
% --- Table 2 (from your markdown) ---
\begin{table}[h]
\centering
\caption{Forecasting results across six geopolitical scenarios. $P(\alpha)$ is
the primary attractor's normalised Bayesian posterior. $c_{\mathrm{final}} =
c_0 \cdot |T|_7$ is the calibrated final confidence under the 7-adic
valuation, where phase transitions occur at steps $T \in \{7, 14, 21,\ldots\}$.
Since $\mathtt{horizon\_steps} = 6 < 7$, all scenarios have $|6|_7 = 1$ (no
domain-level phase transition), so $c_{\mathrm{final}} = c_0$.}
\label{tab:results}
\begin{tabular}{@{}llllll@{}}
\toprule
Scenario & Primary Attractor & $P(\alpha)$ & Secondary Attractor & Bifurcation? & $c_{\mathrm{final}}$ \\
\midrule
US-China tech decoupling & Technology Standards Leadership & 0.51 & Tech Decoupling / Technology Iron Curtain & Step 2 & 0.76 \\
US-China trade war & Hegemonic Sanctions & 0.58 & Trade War / Decoupling & No & 0.74 \\
US-China financial isolation & Hegemonic Sanctions & 0.62 & Financial Reintegration & No & 0.79 \\
China-Taiwan military coercion & Multilateral Alliance Sanctions & 0.54 & Non-State Armed Proxy Conflict & Step 3 & 0.72 \\
China-Taiwan invasion & Multilateral Alliance Sanctions & 0.66 & Non-State Armed Proxy Conflict & No & 0.75 \\
Russia-Ukraine war trajectory & Multilateral Alliance Sanctions & 0.61 & Resource Dependency / Energy Weaponisation & Step 4 & 0.73 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Historical Backtesting Results}

Table~\ref{tab:backtest} reports direction accuracy, Brier scores, and composite
weighted scores across 8 curated historical events. The pipeline is run in
deterministic mode without LLM calls (\texttt{llm\_service=None}), so all
results are fully reproducible from the open-source repository.

\begin{table}[h]
\centering
\caption{Historical backtesting results ($N=8$ curated events with known outcomes). Direction
score: 1.0 = exact pattern match, 0.5 = same ontology domain, 0.0 = miss. Brier score:
$(p - \text{outcome})^2$, lower is better. Attractor score: 1.0 if realized direction in
attractor set, 0.5 if same domain. Composite $= 0.4 \times \text{dir} + 0.3 \times
(1 - \text{Brier}) + 0.3 \times \text{att}$. All runs use \texttt{llm\_service=None}
(deterministic mode); scores computed by \texttt{experiments/backtest\_evaluator.py}.
Confidence priors are ontology-defined; $c_{\mathrm{final}} = c_0$ since all horizons
$T < 7$.}
\label{tab:backtest}
\begin{tabular}{@{}p{0.28\linewidth}p{0.21\linewidth}p{0.06\linewidth}p{0.07\linewidth}p{0.06\linewidth}p{0.09\linewidth}@{}}
\toprule
Event & Realized Direction & Dir & Brier & Att & Comb. \\
\midrule
HK protests 2019         & Great-Power Coercion     & 1.00 & 0.090 & 1.00 & 0.973 \\
Wuhan COVID 2019         & Supply Chain Dependency  & 0.00 & 0.084 & 0.50 & 0.425 \\
Russia-Ukraine 2022      & Interstate Mil.\ Conflict& 1.00 & 0.063 & 1.00 & 0.981 \\
US chip controls 2023    & Entity-List Tech Blockade& 1.00 & 0.040 & 1.00 & 0.988 \\
Huawei ban 2019          & Entity-List Tech Blockade& 1.00 & 0.040 & 1.00 & 0.988 \\
Taiwan drills 2022       & Great-Power Coercion     & 1.00 & 0.090 & 1.00 & 0.973 \\
Brexit withdrawal 2020   & Trade War / Decoupling   & 1.00 & 0.096 & 1.00 & 0.971 \\
SCO expansion 2023       & Formal Military Alliance & 0.00 & 0.073 & 0.50 & 0.428 \\
\midrule
\textbf{Average (8)}     & ---                      & \textbf{0.75} & \textbf{0.072} & \textbf{0.88} & \textbf{0.841} \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textit{Interpretation}: EL-DRUIN achieves exact pattern match (Dir $= 1.0$) on 6
of 8 events---those with a direct ontology-registry counterpart (military conflict, technology
blockade, coercive deterrence, trade decoupling). The two misses (Wuhan COVID supply-chain
disruption, SCO multilateral expansion) reflect events whose dominant trajectory pattern falls
outside the current 21-pattern registry or requires a domain not strongly represented in the
composition rules. The high composite score (0.841 average) is driven by high ontology priors
(0.70--0.80) and the fact that all 8 outcomes were realized (\texttt{outcome\_realized = True}),
meaning the Brier score is small whenever confidence is high. This result honestly indicates
that EL-DRUIN performs well within its ontology coverage and degrades predictably outside it.

\paragraph{On evaluation scope.}
We acknowledge that 14 news excerpts and 8 historical events constitute a
limited evaluation dataset by the standards of large-scale machine learning
benchmarks. EL-DRUIN is not a statistical prediction system: its output is a
deterministic, ontology-grounded symbolic trace, not a learned probability
estimate. The evaluation demonstrates (1) that the symbolic pipeline produces
interpretable, auditable outputs on representative geopolitical scenario types,
and (2) that direction accuracy degrades predictably outside ontology coverage
(the two misses correspond precisely to patterns absent from the 21-pattern
registry). For these demonstration objectives, 22 scenarios spanning nine
thematic categories provide sufficient coverage of the ontology's expressive
range. Scaling to larger benchmarks---including GDELT historical outcomes or
ICEWS event archives---is identified as future work (Section~9), where the
bottleneck is the construction of ground-truth attractor labels rather than
computational cost.

\subsection{Interpretation of Selected Results}

\textbf{US-China Technology Decoupling.}
Starting from \{Hegemonic Sanctions, Entity-List Technology Blockade,
Tech Decoupling / Technology Iron Curtain\}, the composition
(Entity-List Technology Blockade $\oplus$ Tech Decoupling / Technology
Iron Curtain) $\to$ Technology Standards Leadership fires with high
$\mathit{lie\_similarity}$ in Step~1, establishing technology standards
fragmentation as the dominant attractor. A bifurcation at Step~2
($P$ difference $< 0.15$) indicates genuine uncertainty between two
structurally distinct terminal states: a two-standards world (Technology
Standards Leadership) versus a full decoupling regime (Tech Decoupling /
Technology Iron Curtain). This bifurcation is analytically significant:
it corresponds to the empirically contested question of whether the
US-China tech competition produces parallel ecosystems or complete
systemic separation.

\textbf{China-Taiwan Invasion Scenario.}
Starting from \{Interstate Military Conflict, Non-State Armed Proxy
Conflict, Hegemonic Sanctions, Multilateral Alliance Sanctions\}, the
composition (Hegemonic Sanctions $\oplus$ Formal Military Alliance)
$\to$ Multilateral Alliance Sanctions fires immediately via the
composition rule, producing a high-probability ($P(\alpha) \approx 0.66$)
attractor at Multilateral Alliance Sanctions. This result is
substantively interpretable: the system predicts that a Taiwan invasion
scenario converges toward a broad multilateral sanctions coalition
rather than ongoing kinetic conflict---consistent with the historical
pattern following the Russia-Ukraine invasion. The inverse transition
Interstate Military Conflict $\to$ Ceasefire / Peace Agreement appears
as a low-weight ($< 5\%$) path.

\textbf{Confidence Calibration.}
Final confidence values range from 0.72 to 0.79, reflecting the initial
ontology priors (0.68--0.80) multiplied by $|6|_7 = 1$ (since six is not a
multiple of 7, no domain-level phase transition occurs within the six-step
horizon). Under the 7-adic formulation, $c_{\mathrm{final}} = c_0 \cdot |T|_7 = c_0$
for all scenarios with $T < 7$. This contrasts with the previous geometric
decay $\lambda^6 \approx 0.38$: the p-adic approach recognises that a
six-step horizon does not cross a Sylow-7 phase boundary, and therefore
should not decay the confidence below the ontological prior. We consider this
more epistemically honest: decay should be \emph{structurally motivated}, not
uniformly applied.

\paragraph{Extended calibration comparison at $T \in \{7, 14\}$.}
To illustrate the behavioral difference between 7-adic calibration and geometric
decay on longer horizons (beyond the $T=6$ evaluation window), take $c_0 = 0.75$
as the assumed initial confidence and consider:
at $T=7$: $c_{\mathrm{final}}^{\text{7-adic}} = c_0 \cdot |7|_7 = c_0/7 \approx 0.107$,
whereas $c_{\mathrm{final}}^{\lambda} = 0.75 \cdot 0.85^7 \approx 0.268$.
At $T=14$: $c_{\mathrm{final}}^{\text{7-adic}} = c_0/7 \approx 0.107$ (unchanged since
$v_7(14) = 1$), whereas $c_{\mathrm{final}}^{\lambda} \approx 0.095$.
This reveals the fundamental behavioral distinction: geometric decay is
\emph{monotone} (confidence always decreases), while 7-adic calibration is
\emph{structurally punctuated}---flat within a domain-period, then sharply
discounted at multiples of 7 where cross-domain phase transitions are
structurally plausible. The current evaluation uses $T=6$ for all scenarios
(within a single 7-domain period), so both formulas yield $c_{\mathrm{final}} = c_0$.
A multi-week horizon experiment is deferred to future work.

\subsection{Pattern Glossary}

The full glossary of all 21 registered patterns---mapping internal code keys to
English display names and geopolitical domains---is available in the open-source
repository at \texttt{backend/intelligence/pattern\_i18n.py}.
Representative entries are shown in Table~\ref{tab:patterns} (Section~4).

\subsection{Baseline Comparison and Ablation}

Table~\ref{tab:baseline} summarises the qualitative comparison between EL-DRUIN
and the GPT-4o baseline on the properties most relevant to ontology-guided
reasoning: traceability, auditability, stability, and constraint compliance.
Baselines 3 and 4 (rule-only and geometric decay) are described in
Section~\ref{sec:methodology}; their quantitative ablation results on the
news-analysis dataset are available in the repository under
\texttt{experiments/results/}.

\begin{table}[h]
\centering
\caption{System comparison across five configurations. EL-DRUIN (Full) vs
GPT-4o ($N=5$, OpenAI, April 2026) vs ablation variants. $\sigma$ denotes
standard deviation across repeated runs on the same input.}
\label{tab:baseline}
\begin{tabular}{@{}p{0.28\linewidth}p{0.13\linewidth}p{0.13\linewidth}p{0.13\linewidth}p{0.13\linewidth}p{0.13\linewidth}@{}}
\toprule
Metric & EL-DRUIN (Full) & B3: Rule-only & B4: Geom.\ decay & B1: LLM-only (GPT-4o) \\
\midrule
Confidence source & Ontology prior $\times$ Bayes (deterministic) & Same & Same & Self-reported \\
Compute trace & Full & Full & Full & None \\
Verifiable & Yes & Yes & Yes & No \\
Stability $\sigma$ & 0.000 & 0.000 & 0.000 & $>0$ \\
Lie similarity & \checkmark & $\times$ (uniform) & \checkmark & N/A \\
7-adic calibration & \checkmark & $\times$ & $\times$ & N/A \\
Entity invention guard & Enforced & Enforced & Enforced & Not enforced \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Comparison with Statistical Event Coding Systems}
\label{sec:system-comparison}

Table~\ref{tab:system-comparison} extends the comparison to include
GDELT-style statistical event coding systems (GDELT~\cite{leetaru2013gdelt},
ICEWS~\cite{boschee2015icews}, and CAMEO-based pipelines~\cite{schrodt2012cameo}).
These systems are widely deployed for large-scale political event monitoring, but
they differ fundamentally from EL-DRUIN in their output semantics: they aggregate
events statistically and do not produce per-analysis confidence derivations,
provenance records, or trajectory composition chains.

\begin{table}[h]
\centering
\caption{System-level comparison across four property dimensions. ``N/A'' indicates
the property is not applicable or not produced by design; ``---'' indicates the system
does not provide this output type. GDELT/ICEWS values represent the class of
statistical event aggregation systems; EL-DRUIN is the full system.}
\label{tab:system-comparison}
\small
\begin{tabular}{@{}p{0.26\linewidth}p{0.17\linewidth}p{0.17\linewidth}p{0.17\linewidth}p{0.15\linewidth}@{}}
\toprule
Property & GDELT / ICEWS / CAMEO & LLM-only (GPT-4o) & EL-DRUIN (Full) \\
\midrule
\textbf{Output type}
  & Event frequency counts / trend scores
  & Free-text trajectory judgment
  & Structured attractor + posterior \\
\textbf{Confidence source}
  & Implicit (event count / CAMEO score)
  & Self-reported (no derivation)
  & Ontology prior $\times$ Bayesian posterior \\
\textbf{Compute trace / provenance}
  & --- (no per-analysis trace)
  & --- (no trace)
  & Full \texttt{compute\_trace\_ref} per output \\
\textbf{Verifiable outputs}
  & No (aggregate statistics)
  & No (LLM assertion)
  & Yes (deterministic formula) \\
\textbf{Stability $\sigma$}
  & Deterministic (historical DB)
  & $> 0$ (stochastic)
  & 0.000 (deterministic) \\
\textbf{Trajectory composition}
  & No (event counts only)
  & Implicit (unstructured)
  & Yes (rule-based + Lie similarity) \\
\textbf{Training data required}
  & Yes (historical event DB)
  & Yes (LLM pre-training)
  & No (ontology-only) \\
\textbf{Explanation generation}
  & No
  & Yes (unstructured)
  & Yes (constrained, locked fields) \\
\textbf{Real-time news analysis}
  & Partial (batch pipelines)
  & Yes
  & Yes \\
\textbf{Ontology-typed patterns}
  & CAMEO codes (flat vocabulary)
  & None
  & 21 typed patterns, $7 \times 3$ factorization \\
\bottomrule
\end{tabular}
\end{table}

\noindent The key differentiator is not raw predictive accuracy (which would require
large-scale ground-truth comparison beyond the scope of this paper) but rather the
\textbf{auditability and structural grounding} of the outputs. GDELT and ICEWS provide
indispensable large-scale event monitoring, but their confidence metrics are implicit
in event frequencies and cannot be traced to ontology-defined prior beliefs or
composition-rule derivations. EL-DRUIN is designed to complement, not replace, these
systems: a GDELT alert can serve as an input trigger for EL-DRUIN's structured
trajectory analysis.

EL-DRUIN sits at the intersection of four research areas: ontology-based knowledge
graph reasoning, political event coding and geopolitical analysis systems,
explainable neuro-symbolic AI, and LLM-augmented ontology and KG systems.

\subsection{Ontology and Knowledge Graph Reasoning}

Formal ontologies represented in OWL 2~\cite{grau2008owl2,w3c2012owl2} provide the
foundation for semantic web reasoning, enabling inference over typed entities and relations. Large
knowledge graphs such as DBpedia~\cite{bizer2009dbpedia}, Wikidata~\cite{vrandecic2014wikidata},
and YAGO~\cite{suchanek2007yago} demonstrate the scale at which entity-relation triples
can be curated and queried. Hogan et al.~\cite{hogan2021knowledge} provide a
comprehensive survey of knowledge graph construction, representation, and reasoning.
Ji et al.~\cite{ji2021survey} survey representation learning approaches for knowledge
graphs, including embedding methods such as RotatE~\cite{sun2019rotate} that capture
relational structure geometrically.

Temporal knowledge graphs extend static KGs with time-stamped facts and require
reasoning about relational change over time~\cite{trivedi2017know,lacroix2020tensor,li2021temporal}.
Allen's interval algebra~\cite{allen1983maintaining} provides a foundational formalism
for temporal relation reasoning. EL-DRUIN addresses temporal trajectory reasoning
differently: rather than learning embeddings from historical triples, it applies
ontology-registered composition rules to derive forward-projected attractor states,
without requiring labelled training data.

\subsection{Political Event Coding and Geopolitical Analysis Systems}

Computational political science has developed structured event coding schemes to
represent political interactions as typed actor-action-target triples.
CAMEO~\cite{schrodt2012cameo} provides a widely used controlled vocabulary for
political event types. GDELT~\cite{leetaru2013gdelt} and ICEWS~\cite{boschee2015icews}
are large-scale event databases built on CAMEO coding, enabling statistical analysis
of political trends at global scale. These systems aggregate events statistically;
they do not model pattern composition or produce algebraically grounded trajectory
predictions with provenance traces.

Probabilistic graphical models have been applied to conflict
prediction~\cite{ward2010perils}, but these require labelled training data and do
not provide symbolic interpretability of the composition structure. EL-DRUIN requires
no training data; its inference is entirely symbolic and determined by the manually
curated pattern registry, making its reasoning process fully auditable.

\subsection{Explainable Neuro-Symbolic AI}

The neuro-symbolic AI research programme~\cite{besold2017neural,garcez2020neural}
aims to combine the pattern recognition strengths of neural networks with the
interpretability and compositionality of symbolic reasoning. DeepProbLog~\cite{manhaeve2018deepproblog}
integrates probabilistic logic programming with neural networks, enabling
differentiable learning over structured symbolic programs. EL-DRUIN is inspired by
this separation of concerns: symbolic inference governs all numerical outputs while
a neural component (the LLM) handles natural-language rendering only.

Neuro-symbolic concept learning~\cite{mao2019neuro} demonstrates that structured
symbolic programs can constrain neural predictions; EL-DRUIN applies an analogous
principle to geopolitical reasoning, where ontology-defined composition rules
constrain the space of valid trajectory transitions. Our system's constrained LLM verbalization
mechanism extends this idea by structurally prohibiting the LLM from modifying
pre-computed fields, ensuring that explanation fidelity is enforced at the
architecture level rather than by prompt engineering alone.

\subsection{LLMs for Ontology and Knowledge Graph Analysis}

The intersection of LLMs and knowledge graphs is an active research frontier.
Pan et al.~\cite{pan2023unifying} provide a roadmap for unifying LLMs and KGs,
identifying complementary strengths: LLMs contribute linguistic flexibility and
world knowledge, while KGs contribute structured, auditable, and provenance-rich
representations. Agent-OM~\cite{fathallah2023agentom} and
LLMs4OM~\cite{babaei2024llms4om} apply LLM agents to ontology matching, using
syntactic, lexical, and semantic matching layers. EL-DRUIN borrows the multi-layer
entity grounding approach from these works for its pattern activation stage.

Maedche and Staab~\cite{maedche2001ontology} established ontology learning from
text as a research direction; recent work on chain-of-thought prompting~\cite{wei2022chain}
has shown that structured step-by-step reasoning can substantially improve LLM output
quality, motivating the use of structured prompts with locked fields in EL-DRUIN's
verbalization mechanism. In EL-DRUIN, the LLM is not used for ontology
construction or numerical inference, but exclusively for natural-language
verbalization of pre-computed structured results---a deliberate architectural
constraint that separates concerns and ensures reproducibility.

Noy et al.~\cite{noy2019industry} document lessons from deploying industry-scale
knowledge graphs, highlighting the importance of provenance tracking and
schema governance---properties that EL-DRUIN builds in at the ontology layer
through its typed pattern registry and compute trace mechanism.

\section{Discussion}
\label{sec:discussion}

\subsection{What EL-DRUIN Can and Cannot Claim}

EL-DRUIN is a \textbf{system demonstration} paper: its primary contribution is a
working, open-source, ontology-guided reasoning pipeline with demonstrable
properties (determinism, provenance, auditability, constrained LLM verbalization).
It is not a claim that geopolitical trajectories are fully predictable, nor that
the algebraic structures described are the uniquely correct formalization of
geopolitical dynamics.

Specifically:
\begin{itemize}
  \item \textbf{The $7 \times 3$ factorization} is a principled design choice, not
    an empirical discovery. A different ontology designer might choose a different
    partition. The factorization's value lies in its organizational clarity and
    reasoning consequences, not in any claim that geopolitical reality has exactly
    21 fundamental patterns.
  \item \textbf{The non-Archimedean confidence calibration} is a design-motivated
    formalization. We do not claim that geopolitical confidence follows the formal
    theory of p-adic numbers, nor that the composition table constitutes a
    mathematically complete group. The 7-based valuation is a principled
    alternative to the ad hoc constant $\lambda = 0.85$, motivated by the domain
    partition structure.
  \item \textbf{The Lie algebra approximation} provides a continuous-valued
    plausibility score for discrete composition steps, but the 8-dimensional
    vectors are manually annotated and introduce expert subjectivity. The
    commutator structure provides a mathematically well-defined path-independence
    diagnostic; its geopolitical interpretation is heuristic.
  \item \textbf{The backtesting results} use curated ground-truth labels assigned
    by the authors based on historical record. While the events are well-documented
    and the labels are defensible, they are not the product of a blind annotation
    process with inter-annotator agreement measurement.
\end{itemize}

\subsection{Limitations and Future Work}

\textbf{Pattern vector assignment.}
The 8-dimensional vectors $\varphi(p)$ are currently assigned by expert annotation,
introducing potential subjectivity. Future work should learn these vectors from
labelled event data or from comparative political analysis corpora, and measure
inter-annotator agreement on the annotations.

\textbf{Composition table completeness.}
The current composition table contains 16 rules over the registered patterns
(coverage $\approx 3.6\%$ of all possible pairs). The vast majority of compositions
are undefined ($\bot$). Completeness could be improved by inferring additional
rules from historical event co-occurrence data. Until the table is substantially
more complete, forward simulation results should be interpreted as indicative
rather than exhaustive.

\textbf{Ontology curation subjectivity.}
The pattern registry, domain assignments, mechanism classes, and confidence priors
are all manually curated. Different domain experts might curate different registries.
A rigorous ontology evaluation would require independent expert validation of the
21 patterns and their properties.

\textbf{No claim of geopolitical certainty.}
EL-DRUIN produces attractor predictions with calibrated confidence values, but
these should be understood as structured scenario projections from a symbolic
model, not as probabilistic forecasts validated against historical base rates.
The system does not model the full complexity of geopolitical systems; it provides
a structured analytical scaffold.

\textbf{Algebraic completeness.}
As noted in Section~\ref{sec:pattern-algebra}, the composition table does not
constitute a full group structure (only 16 of 441 possible pairs are defined). The
$7 \times 3$ factorization and the Sylow-inspired organizational scheme are best
understood as structured design heuristics. A rigorous algebraic treatment would
require either (a) completing the Cayley table to fully define the group operation
or (b) formally characterizing the partial algebra as a specific algebraic structure
(e.g., a partial groupoid or a semigroup with zero).

\textbf{Evaluation scope.}
The current evaluation covers 14 news excerpts and 8 backtesting events. A more
rigorous evaluation would use a larger, independently curated dataset, involve
domain expert assessment of the structured outputs, and include human evaluation
of explanation quality. We consider this a priority for future work.

\section{Conclusion}

We have presented EL-DRUIN, an ontology-guided reasoning system for
geopolitical trajectory analysis that combines formal ontology, symbolic
composition inference, and constrained LLM explanation generation. The
central contribution is a \textbf{provenance-preserving reasoning pipeline}:
starting from typed event extraction, the system activates ontology-registered
patterns, applies rule-based composition to derive trajectory transitions,
scores transitions via Lie algebra similarity, and computes a fully auditable
Bayesian posterior—converging to attractor states that constitute the
long-run structural prediction. The 21 patterns are organized in a
$7 \times 3$ domain--mechanism factorization that provides a canonical
address for each pattern and grounds confidence calibration in the
ontology structure. The \textbf{Lie algebra embedding}
provides a continuous-valued plausibility weight ($\mathit{lie\_similarity}$)
for each discrete composition step, coupling the discrete algebraic
structure with a continuous geometric one. \textbf{Non-Archimedean confidence calibration}
replaces the ad hoc geometric decay $\lambda^t$ with a 7-based valuation
$|t|_7$, motivated by the seven-domain ontology partition: confidence
remains stable within a domain-aligned horizon and updates sharply at
domain-boundary multiples of 7. \textbf{Structured bifurcation detection}
identifies instability via the 7-adic valuation of the posterior weight
gap, providing a domain-granularity-aligned criterion for near-tie detection. The system is fully open-source, with a working frontend that exposes
all computation traces. We believe this architecture---algebraic
structure as the primary reasoning engine, language model as a
constrained interpreter---represents a promising direction for the next
generation of AI-assisted geopolitical analysis systems.

\paragraph{Positioning.}
We frame the non-Archimedean confidence and ultrametric components as
\emph{design-motivated formalizations}: the formal connection to p-adic geometry
lies beyond the scope of this work, and we make no claim that the partial
composition table constitutes a full group structure. What we implement is a
7-based non-Archimedean calibration on a finite ontology-partitioned pattern
space, providing a principled, domain-sensitive alternative to the ad hoc decay
constant $\lambda = 0.85$. The $7 \times 3$ factorization and the Sylow-inspired
organizational scheme are best understood as a \emph{structured design heuristic}
that makes confidence calibration interpretable and auditable.

\bibliographystyle{plain}
\bibliography{main}

\end{document}